# Results

```{r setup, file = "R/chapter_start.R", include = FALSE}
# a number of commands need to run at the beginning of each chapter. This
# includes loading libraries that I always use, as well as options for 
# displaying numbers and text.
library(mlogit)
library(wesanderson)
```

This section presents results on the nutrition environment in each of the three
communities of Utah County, West Salt Lake County, and San Juan County, along with 
destination choice model estimates and their application to creating accessibility
maps of each community and the entire state of Utah.

## Nutrition Environment

```{r loadnems}
tar_load(nems_groceries)
nems_groceries <- nems_groceries |> 
  ungroup() |> 
  mutate(
    type = ifelse(type == "Trading Post", "Other", type),
    inflation = ifelse(county == "Utah", market * 1.094, market),
    county = factor(county, levels = c("Utah", "Salt Lake", "San Juan"))
  )
```

Though some basic descriptive statistics of the grocery store attributes were 
presented in @tbl-nems, some additional exploration of the these
attributes is valuable to understand the nutrition environment in these three
communities.

```{r nems-market-avail}
#| label: fig-nems-market-avail
#| fig-cap: Relationship between NEMS availability score and market basket score in each study community. Utah county prices adjusted for 2021-2022 annual inflation.
ggplot(nems_groceries, aes(x = availability, y = market, group = county)) + 
  geom_point(alpha = 0.5, mapping = aes(color = type)) + 
  geom_smooth(method = "lm") +
  facet_grid(~county) + 
  theme_bw() + xlab("NEMS Availability Score")  +
  ylab("Market Basket Cost [$]") +
  scale_color_manual("Store Type", values = wesanderson::wes_palette('Darjeeling1')[c(1:3, 5)])
```

@fig-nems-market-avail presents the relationship between the recorded NEMS availability
score and the USDA market basket cost at the stores by community and store type. 
In all three communities, the relationship is strongly negative, with stores that
stock more varieties of goods also having overall lower prices for those goods. 
This is emphasized by the bottom-right quadrants of these plots (high availability, 
low-cost) being dominated by full-service grocery stores, which have more availability
and lower prices than convenience stores or dollar stores, but require higher traffic
and demand to make up for their lower profit margins.
Average prices in Utah County are lower than prices in the other 
two communities across the availability spectrum; this is true even after
adjusting for 9.4% annual inflation between March 2021 and March 2022 in food 
products [@bureauoflaborstatistics2023].

```{r nems-cost-avail}
#| label: fig-nems-cost-avail
#| fig-cap: Relationship between NEMS availability score and cost score in each study community.
ggplot(nems_groceries, aes(x = availability, y = cost)) + 
  geom_point(alpha = 0.5, mapping = aes(color = type)) + 
  geom_smooth(method = "lm") +
  facet_grid(~county) + 
  theme_bw() + xlab("NEMS Availability Score")  +
  ylab("NEMS Cost Score") +
  scale_color_manual("Store Type", values = wesanderson::wes_palette('Darjeeling1')[c(1:3, 5)])
```

@fig-nems-cost-avail shows the relationship between the NEMS availability and cost 
scores. In this case the relationship is generally positive, with stores that 
stock more healthful options also placing these options at competitive prices.
Conversely, stores that have fewer of these options tend to place the options
they do stock at a higher price point. This relationship between availability
and cost of healthful goods is strongest in San Juan County, with convenience
stores anchoring the low-availability, high-premium quadrant for healthy food.
It should be noted that these convenience stores also exist in the Utah County
community, but we explicitly included them in the San Juan data collection as
they are the only food markets of any kind in multiple towns, with dozens of miles
separating towns from each other.


## Destination Choice 

The models were estimated using the mlogit package for R [@mlogit]. To illustrate
the role of different 

```{r loadmodels}
tar_load(sl_models)
tar_load(ut_models)
tar_load(sj_models)
```

```{r model-setups, include=FALSE}
glance_custom.mlogit <- function(x, ...){
  broom:::glance.mlogit(x)
}
f <- function(x) format(round(x, 3), big.mark=",")
gm <- list(
  list("raw" = "AIC", "clean" = "AIC", "fmt" = f),
  list("raw" = "rho20", "clean" = "$\\rho^2_0$", "fmt" = f))

cm <- c(
  "duration_CAR" = "Automobile Travel Time",
  "mclogsum" = " Mode Choice Logsum", 
  "availability" = "NEMS Availability Score",
  "cost" = "NEMS Cost Score",
  "market" = "USDA Market Basket",
  "registers" = "Registers",  
  "typeDollar Store" = "Store Type: Dollar Store", 
  "typeConvenience Store" = "Store Type: Convenience Store",
  "typeOther" = "Store Type: Other"
)
```


```{r sl-models}
#| label: tbl-sl-models
#| tbl-cap: Estimated Models of Salt Lake County
modelsummary(sl_models, gof_map = gm, coef_map = cm)
```


```{r utah-models}
#| label: tbl-utah-models
#| tbl-cap: Estimated Models of Utah County
modelsummary(ut_models, gof_map = gm, coef_map = cm)
```


```{r sl-models}
#| label: tbl-sj-models
#| tbl-cap: Estimated Models of San Juan County
modelsummary(sj_models, gof_map = gm, coef_map = cm)
```



```{r all-comp}
#| label: fig-all-comp
#| fig-cap: Comparison of county model coefficient estimates.
list(
  "Salt Lake" = sl_models[["All"]],
  "Utah"      = ut_models[["All"]],
  "San Juan"  = sj_models[["All"]]
) |> 
  modelplot(coef_map = cm) +
  scale_color_manual(values = wesanderson::wes_palette('Darjeeling1'))
```

## Accessibility

### Statewide Accessibility
