# Results {#sec-results}

```{r setup, file = "R/chapter_start.R", include = FALSE, cache = FALSE}
# a number of commands need to run at the beginning of each chapter. This
# includes loading libraries that I always use, as well as options for 
# displaying numbers and text.
library(mlogit)
library(wesanderson)
library(sf)
library(ggspatial)
library(viridisLite)
```

This section presents results on the nutrition environment in each of the three
communities of Utah County, West Salt Lake County, and San Juan County, along
with destination choice model estimates and their application to creating
accessibility maps of each community and the entire state of Utah.

## Nutrition Environment {#sec-nems}

```{r loadnems}
tar_load(nems_groceries)
nems_groceries <- nems_groceries |> 
  ungroup() |> 
  mutate(
    type = ifelse(type == "Trading Post", "Other", type),
    inflation = ifelse(county == "Utah", market * 1.094, market),
    county = factor(county, levels = c("Utah", "Salt Lake", "San Juan"))
  )
```

Though some basic descriptive statistics of the grocery store attributes were
presented in @tbl-nems, some additional exploration of these attributes is
valuable to understand the nutrition environment in these three communities.

```{r nems-market-avail}
#| label: fig-nems-market-avail
#| fig-cap: Relationship between NEMS availability score and market basket score in each study community. Utah county prices adjusted for 2021-2022 annual inflation.
ggplot(nems_groceries, aes(x = availability, y = market, group = county)) + 
  geom_point(alpha = 0.5, mapping = aes(color = type)) + 
  geom_smooth(method = "lm") +
  facet_grid(~county) + 
  theme_bw() + xlab("NEMS Availability Score")  +
  ylab("Market Basket Cost [$]") +
  scale_color_manual("Store Type", values = wesanderson::wes_palette('Darjeeling1')[c(1:3, 5)])
```

@fig-nems-market-avail presents the relationship between the recorded NEMS
availability score and the USDA market basket cost at the stores by community
and store type. In all three communities, the relationship is strongly negative,
with stores that stock more varieties of goods also having overall lower prices
for those goods. This is emphasized by the bottom-right quadrants of these plots
(high availability, low-cost) being dominated by full-service grocery stores,
which have more availability and lower prices than convenience stores or dollar
stores, but require higher traffic and demand to make up for their lower profit
margins. Average prices in Utah County are lower than prices in the other two
communities across the availability spectrum; this is true even after adjusting
for 9.4% annual inflation between March 2021 and March 2022 in food products
[@bureauoflaborstatistics2023].

```{r nems-cost-avail}
#| label: fig-nems-cost-avail
#| fig-cap: Relationship between NEMS availability score and cost score in each study community.
ggplot(nems_groceries, aes(x = availability, y = cost)) + 
  geom_point(alpha = 0.5, mapping = aes(color = type)) + 
  geom_smooth(method = "lm") +
  facet_grid(~county) + 
  theme_bw() + xlab("NEMS Availability Score")  +
  ylab("NEMS Cost Score") +
  scale_color_manual("Store Type", values = wesanderson::wes_palette('Darjeeling1')[c(1:3, 5)])
```

@fig-nems-cost-avail shows the relationship between the NEMS availability and
cost scores. In this case the relationship is generally positive, with stores
that stock more healthful options also placing these options at competitive
prices. Conversely, stores with fewer options tend to place the options they do
stock at a higher price point. This relationship between availability and cost
of healthful goods is strongest in San Juan County, with convenience stores
anchoring the low-availability, high-premium quadrant for healthy food. It
should be noted that these convenience stores also exist in the Utah County
community, but we explicitly included them in the San Juan data collection as
they are the only food markets of any kind in multiple towns, with dozens of
miles separating towns from each other.

## Destination Choice {#sec-estimation}

Using the data collected and MNL destination choice model as described in
@sec-methods, we estimate a series of model specifications in each community
with the `mlogit` package for R [@mlogit]. To illustrate the role of different
data elements on destination choice, we develop and estimate four different
utility equations: 
\begin{align*}
\mathrm{Access} &= \beta_{MCLS}( k_{ij})\\
\mathrm{NEMS} &= \beta_{n-a} (\mathrm{NEMS-Availability}) + \beta_{n-c}\mathrm({NEMS-Cost})\\
\mathrm{Attributes} &= \beta_{mkt} (\mathrm{Market Basket}) + \mathbf{\beta}_{type}(\mathrm{Type})\\
\mathrm{All} &= \mathrm{Access} + \mathrm{NEMS} + \mathrm{Attributes}\\
\end{align*} 
The Access model includes only the mode choice logsum described in
@eq-mcls. The NEMS model includes the NEMS cost and availability scores
describing the goods the store offers, while the Attributes model contains
information that might be more conventionally available to shoppers including
the size, type, and average prices at the store. As the nutrition environment in
each community contains different types of stores, the specific type
coefficients differ by community. The All model contains all of the other three
sets of estimated coefficients.

```{r loadmodels}
tar_load(sl_models)
tar_load(ut_models)
tar_load(sj_models)
```

```{r model-setups, include=FALSE}
glance_custom.mlogit <- function(x, ...){
  broom:::glance.mlogit(x)
}
f <- function(x) format(round(x, 3), big.mark=",")
gm <- list(
  list("raw" = "AIC", "clean" = "AIC", "fmt" = f),
  list("raw" = "rho20", "clean" = "$\\rho^2_0$", "fmt" = f))

cm <- c(
  "duration_CAR" = "Automobile Travel Time",
  "mclogsum" = " Mode Choice Log-sum", 
  "availability" = "NEMS Availability Score",
  "cost" = "NEMS Cost Score",
  "market" = "USDA Market Basket",
  "total_registers" = "Registers",  
  "typeDollar Store" = "Store Type: Dollar Store", 
  "typeConvenience Store" = "Store Type: Convenience Store",
  "typeOther" = "Store Type: Other"
)
```

```{r utah-models}
#| label: tbl-utah-models
#| tbl-cap: Estimated Models of Utah County
modelsummary(ut_models, gof_map = gm, coef_map = cm, 
             stars = c('*' = 0.05, '**' = 0.01),
             statistic = "({statistic})",
             notes = list("t-statistics in parentheses"))
```

@tbl-utah-models presents the estimated coefficients in the Utah County
community. In general, the utility coefficients are statistically significant
and in a direction that would be expected by informed hypothesis. The Access
model has a positive coefficient on its mode choice log-sum term, which
indicates that as the mode choice logsum between a block group and a store
increases --- indicating lower travel costs between Census block groups and the
store, because travel times in @eq-mcls have a negative relationship with
utility --- a higher proportion of mobile devices residing in that block group
are observed to travel to that store. The NEMS model shows a positive
relationship between both environment variables and utility, indicating that
people are more likely to choose stores with higher availability of healthy
goods and more advantageous prices for those goods, all else equal. The
Attributes model suggests that
people are less willing to visit stores with higher prices, fewer registers, and
convenience stores or other non-standard grocery stores with the exception of
dollar stores, which they are *more* attracted to. Combining all of these
variables in the All model retain the significance, direction, and basic scale
of all previous estimates with the exception of the NEMS availability variable.
In this case, it seems that the previous positive relationship may have been a
result of correlation between NEMS availability and other variables such as cost
or the number of registers. And when controlling for all other variables, the
role of transportation access becomes somewhat more important than considering
only distance alone, implying that people are willing to travel somewhat further
for stores with attributes they value.

The overall fit of the four models in @tbl-utah-models is also revealing: the
model with only NEMS variables against almost no predictive power over randomly
selecting any store in the community (as revealed by the $\rho_0^2$ statistic).
Though all sets of variables contribute to the overall fit, it is apparent that
the bulk of model explanatory power is due to transportation proximity.

```{r sl-models}
#| label: tbl-sl-models
#| tbl-cap: Estimated Models of West Salt Lake Valley
modelsummary(sl_models, gof_map = gm, coef_map = cm,
             stars = c('*' = 0.05, '**' = 0.01),
             statistic = "({statistic})",
             notes = list("t-statistics in parentheses"))
```

```{r sj-models}
#| label: tbl-sj-models
#| tbl-cap: Estimated Models of San Juan County
modelsummary(sj_models, gof_map = gm, coef_map = cm,
             stars = c('*' = 0.05, '**' = 0.01),
             statistic = "({statistic})",
             notes = list("t-statistics in parentheses"))
```

@tbl-sl-models presents the estimated coefficients in the west Salt Lake County
community, and @tbl-sj-models presents the estimated coefficients in San Juan
County. The same general story about coefficient direction and hypotheses
applies in both of these communities, except in regards to the NEMS variables.
In Salt Lake, the NEMS cost score appears negative when estimated alone but
becomes positive when other variables are included. In San Juan, these variables
are consistently positive. Additionally, the story of model fit is reversed: in
both Salt Lake and San Juan, the attributes of the store explain more of the
model fit than the transportation impedance term.

```{r all-comp}
#| label: fig-all-comp
#| fig-cap: Comparison of county model coefficient estimates. Absolute square root transformation applied to improve visibility.
#| out-width: "6in"

# use a square root transformation, but have it work in both directions.
absqrt <- scales::trans_new(
  "absqrt",
  function(x){sign(x) * sqrt(abs(x))},
  function(x){sign(x) * x^2},
  breaks = scales::extended_breaks(),
  minor_breaks = scales::regular_minor_breaks(),
  format = scales::format_format(),
  domain = c(-Inf, Inf)
)
list(
  "Utah County"      = ut_models[["All"]],
  "Salt Lake"        = sl_models[["All"]],
  "San Juan"         = sj_models[["All"]]
) |> 
  modelplot(coef_map = cm) +
  scale_x_continuous(trans = absqrt, breaks = c(-1, 0, 1, 3, 5, 10)) +
  scale_color_manual(values = wesanderson::wes_palette('Darjeeling1'))
```

To better visualize how the preferences in the three communities differ from
each other, @fig-all-comp plots the coefficient estimates from the All model in
each community. The mode choice log-sum is strongly significant in all three
communities, but it has its smallest value in San Juan County where people often
must travel long distances to reach any stores. The highest mode choice log-sum
value is in Salt Lake, but this explains a smaller proportion of the model
outcomes than the lower value in Utah County; a possible hypothesis for this
observation may include the higher density of stores in Salt Lake --- attributes
are more important when so many stores are close together --- paired with the
somewhat lower vehicle ownership in that community driving up the coefficient
value. Most other estimates are somewhat consistent across counties, with the
exception of the NEMS variables discussed previously and the role of dollar
stores and other stores in each community.

## Accessibility {#sec-access}

```{r utbgaccess}
tar_load(utbgaccess)
```

With the models estimated in @sec-estimation, we can evaluate the spatial access
of each community. @fig-map-ut shows the value of the grocery store destination
choice log-sum for block groups in Utah County. Unsurprisingly, the block groups
in the core of the urban areas of the region have the highest access to grocery
stores, because this is where the stores are located and also where the
transportation access to multiple destinations is highest. This map also
contains somewhat interesting implications for the equity of access. A perhaps
unique feature of Utah County's demographic geography is that the wealthiest
neighborhoods tend to be located on the mountain benches east of the main urban
areas. This means that in Utah County, at least, the neighborhoods with the
lowest access to grocery stores are actually some of the wealthiest
neighborhoods with the lowest concentrations of ethnic minorities in the region.

```{r map-ut}
#| label: fig-map-ut
#| fig-cap: Modeled access to grocery stores in Utah County
#| out-width: "6in"
pal <- wesanderson::wes_palette("Zissou1", 100, type = "continuous")

ggplot(utbgaccess, aes(fill = access)) +
  ggspatial::annotation_map_tile("cartolight", zoom = 11) +
  geom_sf(alpha = 0.7, lwd = 0) +
    theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_gradientn("Access", colours = pal)
```

Of course, much of this high access in the urban core of Utah County is achieved
by cheap and available automobile transportation. We can consider what access
looks like for those without cars by re-computing the mode choice log-sum
described in @sec-mcls between all block group / store pairs but eliding the
automobile mode, and examining the resulting impact to destination choice
utility. @fig-map-nocar shows the results of this analysis: whereas the total
access (with car included) is a smooth gradient across the valley, the access
for individuals without vehicles is blocky and discontinuous, with neighborhoods
of relatively good access immediately next to neighborhoods with bad or
non-existent access. This may reflect the discontinuous nature of active
transport and public transit facilities in the region, as well as the
auto-dominated locations of many grocery stores. Note also that even for
neighborhoods of relatively good non-vehicle access, the destination choice
log-sum value is substantially lower than the logsum with vehicle access; the
minimum value on with vehicles is just below 0, whereas the *maximum* log-sum
without vehicles is around -100. Because the log-sum occurs on the same scale in
both cases, this represents a serious additional cost for non-vehicle users.

```{r map-nocar}
#| label: fig-map-nocar
#| fig-env: sidewaysfigure
#| fig-cap: "Access to groceries in Salt Lake County with and without a vehicle."
#| fig-subcap: 
#|   - "With vehicles"
#|   - "Without vehicles"
#| layout-ncol: 2
#| out-width: "4in"
tar_load(nocaraccess)

ggplot(nocaraccess, aes(fill = dclogsum)) +
  ggspatial::annotation_map_tile(type = "cartolight", zoom = 12) +
  geom_sf(lwd = 0, alpha = 0.7) +
  scale_x_continuous(limits = c(-112.11, -111.7)) + 
  scale_y_continuous(limits = c(40.5, 40.8)) + 
    theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_gradientn("Access", colours = pal)

ggplot(nocaraccess |> filter(!is.na(nocarlogsum)), aes(fill = nocarlogsum)) +
  ggspatial::annotation_map_tile(type = "cartolight", zoom = 12) +
  geom_sf(lwd = 0, alpha = 0.7) +
  scale_x_continuous(limits = c(-112.11, -111.7)) + 
  scale_y_continuous(limits = c(40.5, 40.8)) + 
    theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_gradientn("Access", colours = pal)
```
