# Application {#sec-scenarios}

```{r setup, file = "R/chapter_start.R", include = FALSE, cache = FALSE}
# a number of commands need to run at the beginning of each chapter. This
# includes loading libraries that I always use, as well as options for 
# displaying numbers and text.
```

```{r setup2, include=FALSE, cache = FALSE}
library(mlogit)
library(wesanderson)
library(sf)
library(ggspatial)
library(viridisLite)
knitr::opts_chunk$set(cache = FALSE)
pal <- wesanderson::wes_palette("Zissou1", 100, type = "continuous")
```

In this section, we develop a series of scenarios to which we apply the models
estimated in @sec-results. These scenarios are constructed to illustrate 
potential applications of a utility-based model, and not as an evaluation
of proposed policies. 
We first describe how each scenario was constructed, and then discuss the
results together.

The accessibility of each scenario is determined using the approximated
utility coefficients of the All model estimated in @sec-results, 

$$
U_{ij}= \beta_{MCLS}( k_{ij}) +  \beta_{n-a} (\mathrm{NEMS-Availability}) +
  \beta_{n-c}\mathrm({NEMS-Cost}) + \\ \beta_{mkt} (\mathrm{Market Basket}) + \pmb{\beta}_{type}(\mathrm{Type})  
$$ {#eq-scenu} 

and the total access benefit is defined as the consumer surplus
given in @eq-cs. Note that this benefit is denominated in units of utility
[@ben-akiva1985], and the coefficients in @eq-scenu serve to convert between the
units of the variable and utility. Specifically, the $\beta_{mkt}$ coefficient
represents how many dollars of grocery cost a person is willing to spend to
increase their utility by one unit.
In actuality, the utility formula in @eq-scenu is a relative utility added by an
unknown constant, $U_{ij} = f(\mathbf{\beta}, X_{ij}) + C$, but this $C$ term is
included in all the alternatives and therefore cancels out in estimation [@train2009].
This means we cannot assess the *absolute* value of utility, but we
can assess the *relative* monetary benefit of the difference between two
scenarios as 
$$
\mathrm{Benefit} = \sum_{i}\left(-\frac{\omega_i}{\beta_{mkt}}(CS_i' - CS_i)\right)
$$ 
with the consumer surplus of the "improved" scenario in each origin zone $i$ 
indicated as $CS_i'$ and the unimproved counterpart as $CS_i$, a weight $\omega$
accounts for the population in the zone, and the $\beta_{mkt}$ converts the
difference in utility into a dollar amount.

## Scenario Descriptions

There are three general strategies we develop scenarios around:

1.  Erect a new grocery store in the community, in a place where one does not
    already exist.
2.  Improve an existing convenience store or dollar store so that it has the
    attributes of a full-service grocery store.
3.  Improve the transit and non-motorized access to stores in a region.
4.  Increase the availability of grocery delivery in the region.

We implement all three strategies in scenarios on the west Salt Lake valley
community. 

### Erect a new store

This strategy assumes that the nutrition environment would benefit from a new
store located in a place that presently has low grocery store access. To examine
the potential for this strategy to improve access to nutrition in each
community, we calculate the change in destination choice log-sum when a new
store is added to the region in a location where access is currently poor. The
new store is a full-service grocery store with a number of registers equal to
the mean of other grocery stores in the community, and NEMS availability score,
NEMS cost score, and market basket cost equivalent to the 75$^{th}$ percentile
for the community. Thus the store is expected to be better-than-average quality
as perceived by the residents of the community. The location for this new store
is at 4100 S and 2700 W in West Valley City.

### Improve an existing store

This strategy assumes that existing stores are in locations that the community
values and can access, but that those stores may not have high availability of
quality goods. To examine the potential for this strategy to improve access to
nutrition, we improve the attributes of an existing dollar store in the
community so that it has the size, prices, and availability of goods as a
full-service grocery store. As above, we create a full-service grocery store
with a number of registers equal to the mean of other grocery stores in the
community, and NEMS availability score, NEMS cost score, and market basket cost
equivalent to the 75$^{th}$ percentile for the community. Thus the store is
expected to be better-than-average quality as perceived by the residents of the
community; the difference from the previous scenario is that the improved store
takes the place of an existing convenience store or dollar store. In this scenario,
we improve the attributes of an existing ethnic store near 2700 W 3500 S in West Valley City.

### Improve transit and non-motorized transport

This strategy assumes that people cannot easily travel to existing stores
because they cannot or do not drive for a variety of reasons, and that the
public and active transport networks provide an insufficient level of
service. To examine the potential for this strategy to improve access to
nutrition, we improve the travel time costs in the Salt Lake community for
non-motorized and public transportation in the region and calculate the change
in destination choice log-sum.

For active transportation, the lack of pedestrian facilities across and
alongside roads both in reality and in the OpenStreetMap dataset may
substantially increase measured walk distances and times. In this scenario, we
replace the times measured from OpenStreetMap using R5 with an idealized
distance function, 
$$ 
t_{\mathrm{walk}} = \frac{\sqrt{2} * d'_{ij}}{v_{\mathrm{walk}}}
$$ {#eq-newwalk} 
where $d'_{ij}$ is the Euclidean (straight-line) distance
between $i$ and $j$ and $v_{ij}$ is an average walking speed equivalent to 3.5
feet per second [@fitzpatrick2006]. The distance is multiplied by the square
root of 2 to reflect the Manhattan distance (along a gridded street system). We
retain the cap on walking distance at 10 kilometers. Though this distance may
radically understate the real walking distance, we are trying to create an
idealized scenario of effectively frictionless active transport. For public
transit, we assume that the frequency of service is such that all transfer and
initial wait times are at most 5 minutes, and that no person must walk more than
10 minutes to access their first public transport service. Presently, the
*mean* walk access time in the scenario region is over 20 minutes, and the
wait time over 10 minutes.

Travel times are improved in this way for all block group --- store pairs in the
west Salt Lake Valley community.

### Grocery delivery services

This strategy assumes that the most effective way to bring quality nutrition to an 
inaccessible location is not to change the distribution of stores or the quality of transport, but to
support and enable delivery services that will bring the nutrition to individuals.
To model this scenario, we assume that 20% of the full-service grocery stores in 
the west Salt Lake Valley community offer a delivery service. This raises the
price of the market basket by $10, but reduces the travel time on all modes to 0
if the store is within 3 kilometers.
All of these assumptions are imperfect: a local grocery chain presently offers
services at a $7 delivery fee, which is not paid by all customers; additionally, 
delivery services require that customers expend time placing online orders; and 
service providers in the region typically place a long-distance charge rather
than a service area limit. Regardless, these assumptions will work for the
purpose of this exercise.

## Scenario Results

```{r loadbase}
tar_load(bg)
tar_load(bg_acs)
sl_base <- tar_read(sl_access)
tar_load(mkt_betas)
```

Using the methodology described above, we recalculated the destination choice
log-sum value for each block group under each scenario, and compared the change
in accessibility resulting from the improvement.

```{r}
#| label: fig-scenario-results
#| fig-env: sidewaysfigure
#| fig-cap: Estimated per-household benefits under the four Salt Lake Valley scenarios.
#| fig-subcap:
#|   - "New full-service store"
#|   - "Improved existing store"
#|   - "Improved non-motorized and transit access"
#|   - "Expanded grocery delivery"
#| layout-ncol: 2
#| out-width: "4in"
s1_altd <- tar_read(s1_access)
tar_load(s1_stores)
s2_altd <- tar_read(s2_access)
tar_load(s2_stores)
s3_altd <- tar_read(s3_access)
tar_load(gtfs_shape)
s4_altd <- tar_read(s4_access)
tar_load(s4_stores)

s1cost <- bg |>  select(id = GEOID) |> filter(substr(id, 3, 5) == "035") |> 
  left_join(bg_acs, by = c("id" = "geoid")) |> 
  sf::st_transform(4326) |> 
  left_join(sl_base |> rename(base = dclogsum)) |> 
  left_join(s1_altd |> rename(alt = dclogsum)) |> 
  mutate(diff = (base - alt) / mkt_betas["SaltLake.market"])

s2cost <- bg |>  select(id = GEOID) |> filter(substr(id, 3, 5) == "035") |>
  left_join(bg_acs, by = c("id" = "geoid")) |>
  sf::st_transform(4326) |>
  left_join(sl_base |> rename(base = dclogsum)) |>
  left_join(s2_altd |> rename(alt = dclogsum)) |>
  mutate(diff = (base - alt) / mkt_betas["SaltLake.market"])

s3cost <- bg |>  select(id = GEOID) |> filter(substr(id, 3, 5) == "035") |>
  left_join(bg_acs, by = c("id" = "geoid")) |>
  sf::st_transform(4326) |>
  left_join(sl_base |> rename(base = dclogsum)) |>
  left_join(s3_altd |> rename(alt = dclogsum)) |>
  mutate(diff = (base - alt) / mkt_betas["SaltLake.market"])

gtfs_shape <- sf::st_simplify(gtfs_shape, dTolerance = 10)

s4cost <- bg |>  select(id = GEOID) |> filter(substr(id, 3, 5) == "035") |>
  left_join(bg_acs, by = c("id" = "geoid")) |>
  sf::st_transform(4326) |>
  left_join(sl_base |> rename(base = dclogsum)) |>
  left_join(s4_altd |> rename(alt = dclogsum)) |>
  mutate(diff = (base - alt) / mkt_betas["SaltLake.market"])

scenario_map_theme <- theme(
  axis.text.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.text.y = element_blank(),
  axis.ticks.y = element_blank()
)

plot_s1 <- ggplot() +
  geom_sf(data = s1cost, aes(fill = diff), lwd = 0) + 
  ggspatial::annotation_map_tile(type = "cartolight", zoom = 11, alpha = 0.6) +
  ggspatial::annotation_north_arrow(style = north_arrow_minimal, location = "tl") +
  ggspatial::annotation_scale() +
  geom_sf(data = s1_stores, color = "black", shape = 21, size = 3, fill = "white") +
  scale_x_continuous(limits = c(-112.11, -111.7)) + 
  scale_y_continuous(limits = c(40.5, 40.8)) + 
  scenario_map_theme +
  scale_fill_viridis_c("Benefit", direction = -1, na.value = NA)

plot_s2 <- ggplot() +
  geom_sf(data = s2cost, aes(fill = diff), lwd = 0) +
  ggspatial::annotation_map_tile(type = "cartolight", zoom = 11, alpha = 0.6) +
  ggspatial::annotation_north_arrow(style = north_arrow_minimal, location = "tl") +
  ggspatial::annotation_scale() +  
  scale_x_continuous(limits = c(-112.11, -111.7)) + 
  scale_y_continuous(limits = c(40.5, 40.8)) + 
  geom_sf(data = s2_stores, color = "black", shape = 21, size = 3, fill = "white") +
  scenario_map_theme +
  scale_fill_viridis_c("Benefit", direction = -1, na.value = NA)

plot_s3 <- ggplot() +
  geom_sf(data = s3cost, aes(fill = diff), lwd = 0) +
  ggspatial::annotation_map_tile("cartolight", zoom = 10, alpha = 0.6) +
  ggspatial::annotation_north_arrow(style = north_arrow_minimal, location = "tl") +
  ggspatial::annotation_scale() +
  geom_sf(data = gtfs_shape |> filter(mode != "BRT"), aes(color = mode)) +
  scale_x_continuous(limits = c(-112.11, -111.7)) + 
  scale_y_continuous(limits = c(40.5, 40.8)) + 
  scenario_map_theme +
  scale_fill_viridis_c("Benefit", direction = -1, na.value = NA) +
  scale_color_manual("Transit", values = c("gray80", "black", "gray50"))

plot_s4 <- ggplot() +
  geom_sf(data = s4cost, aes(fill = diff), lwd = 0) +
  ggspatial::annotation_map_tile("cartolight", zoom = 10, alpha = 0.6) +
  ggspatial::annotation_north_arrow(style = north_arrow_minimal, location = "tl") +
  ggspatial::annotation_scale() +
  geom_sf(data = s4_stores, color = "black", shape = 21, size = 3, fill = "white") +
  scale_x_continuous(limits = c(-112.11, -111.7)) + 
  scale_y_continuous(limits = c(40.5, 40.8)) + 
  scenario_map_theme +
  scale_fill_viridis_c("Benefit", direction = -1, na.value = NA)

plot_s1
plot_s2
plot_s3
plot_s4
```

@fig-scenario-results shows the geographic distribution of benefits associated
with each Salt Lake Valley scenario. The benefits from locating a new store at a
site in the Salt Lake community (@fig-scenario-results-1) are largest
immediately next to the new store, where they exceed
$`r floor(max(s1cost$diff, na.rm = TRUE))`$ for each household each time the
household makes a trip to a grocery store.

The improved-store scenario (@fig-scenario-results-2) is similar in shape, but
the scale of the benefits are not as substantial (a maximum
per-household-trip benefit of less than \$1), and seem to not cover quite as
large a geographic region. The transit and non-motorized access scenario
(@fig-scenario-results-3) spreads smaller benefits over a larger area, while the
grocery-delivery scenario (@fig-scenario-results-4) produces the largest and
most broadly distributed gains.



```{r}
#| label: tbl-scenarios
#| tbl-cap: Scenario Benefits
these_betas <- mkt_betas["SaltLake.market"]


df <- list(
  "New Store" = s1cost,
  "Improved Store" = s2cost,
  "Improved Transport" = s3cost,
  "Delivery Services" = s4cost
) |> 
  bind_rows(.id = "Scenario") |> 
  sf::st_set_geometry(NULL) |> 
  mutate(
    hh = households * diff,
    nw = (100 - white)/100 * households * diff,
    li = households * lowincome/100 * diff,
    zv = households * zero_vehicle / 100 * diff
  ) |> 
  summarise(
    across(hh:zv, ~sum(.x, na.rm = TRUE)),
    .by = Scenario
  )  |> 
  mutate(
    betas = these_betas * -1,
    across(hh:zv, ~scales::dollar(./betas))
  ) |> 
  select(-betas)

if(!knitr::pandoc_to("docx")){
  kbl(df, booktabs = TRUE, col.names = c("Scenario", "Households", "Non-white", "Low-Income", "Zero-Vehicle"),
      digits = 0) |> 
  add_header_above(c(" " = 1, "Weighted by" = 4)) |> 
  kable_styling()
}
```

Although comparing the geographic distribution of benefits is helpful, the
aggregate benefit is more likely to guide policy. Additionally, the aggregate
benefits can be weighted in different ways to understand the effects of the
various policies on different populations. @tbl-scenarios presents the aggregate
benefit from each of the four scenarios 
The Households column multiplies the difference in
destination choice log-sum at each block group by the number of households in
that block group, while the non-white and low-income columns weight the
difference by the share of non-white individuals and low-income 
and zero-vehicle zero-vehicle respectively. All demographic data comes from the
American Community Survey (ACS) 5-year aggregations; the ACS discloses households
by vehicles available at the tract level, while all other household characteristics 
are at the more spatially refined block group level.

These alternate weighting schemes help to illustrate the potential equity of the
benefit distribution should each scenario be pursued. The new store scenario in
the Salt Lake community, for example, has a total benefit of approximately \$13
million. Of this amount, somewhat less than half the benefits go to non-white
individuals and less than one-fifth to low-income households. These ratios are
more or less the same for the store improvement scenario and the improved
transport scenario in the Salt Lake Valley. 

Overall, the improved store brings more than twice the benefits of improving
non-automobile transportation, and the new store more than five times the
benefits. Understanding the costs of these various alternatives is outside the
scope of this research, but the level of infrastructure investment required to
increase transportation level of service to that constructed in the simulation
is likely an order of magnitude higher than the cost of a single new grocery
store. It is acknowledged that improving non-automobile
infrastructure and services would have benefits for all trip types --- beyond just the grocery store trips ---
that we do not attempt to enumerate here. 
It is also not clear whether the level
of improvement simulated in the second scenario could be accomplished within the
envelope of the existing store; it may be that such improvements would meet or
exceed the cost of building a new store on a new site, along with stocking,
staffing, and operating the store.

By far the highest benefit observed in these scenarios is the increased availability
of delivery services, but this scenario must be treated with some skepticism. 
Delivery services place costs on shoppers that are not well-quantified in the existing model:
potential increases in shopping time, uncertainty over selection, inability to
inspect goods prior to purchase, delays in delivery, and so forth. Additionally,
low extant delivery prices may be a function of predatory pricing strategies by
application developers seeking to capture market share first, and achieve
profitability later [@moore2022]. Resolving all of these issues cannot be 
handled in the present analysis, but it must be at least considered that grocery
delivery may have the greatest consumer surplus of these four strategies by a
wide margin.
