# Application {#sec-scenarios}

```{r setup, file = "R/chapter_start.R", include = FALSE, cache = FALSE}
# a number of commands need to run at the beginning of each chapter. This
# includes loading libraries that I always use, as well as options for 
# displaying numbers and text.
```

```{r setup2, include=FALSE, cache = FALSE}
library(mlogit)
library(wesanderson)
library(sf)
library(ggspatial)
library(viridisLite)
```

In this section, we develop a series of scenarios to which we apply the models
estimated in @sec-results. These scenarios are constructed to ascertain what may
be the best strategy to improve nutrition access in a community. We first
describe how each scenario was constructed, and then discuss the results
together.

The accessibility of each scenario is determined using the approximated
utility coefficients of the All model estimated in @sec-results, 
$$
U_{ij}= \beta_{MCLS}( k_{ij}) +  \beta_{n-a} (\mathrm{NEMS-Availability}) +
  \beta_{n-c}\mathrm({NEMS-Cost}) + \\ \beta_{mkt} (\mathrm{Market Basket}) + \pmb{\beta}_{type}(\mathrm{Type})  
$$ {#eq-scenu} 
and the total access benefit is defined as the consumer surplus
given in @eq-cs. Note that this benefit is denominated in units of utility
[@ben-akiva1985], and the coefficients in @eq-scenu serve to convert between the
units of the variable and utility. Specifically, the $\beta_{mkt}$ coefficient
represents how many dollars of grocery cost a person is willing to spend to
increase their utility by one unit.

In actuality, the utility formula in @eq-scenu is a relative utility added by an
unknown constant, $U_{ij} = f(\mathbf{\beta}, X_{ij}) + C$, but this $C$ term is
included in all the alternatives and therefore cancels out [@train2009] in
estimation. This means we cannot assess the *absolute* value of utility, but we
can assess the *relative* monetary benefit of the difference between two
scenarios as $$
\mathrm{Benefit} = \sum_{i}\left(-\frac{\omega_i}{\beta_{mkt}}(CS_i^1 - CS_i^1)\right)
$$ where a weight $\omega$ accounts for the population in each origin zone $i$
and the $\beta_{mkt}$ converts the difference in utility into a dollar amount.

## Scenario Descriptions

There are three general strategies we develop scenarios around:

1.  Erect a new grocery store in the community, in a place where one does not
    already exist.
2.  Improve an existing convenience store or dollar store so that it has the
    attributes of a full-service grocery store.
3.  Improve the transit and non-motorized access to stores in a region.
4.  Increase the availability of grocery delivery in the region.

We implement all three strategies in scenarios on the west Salt Lake valley
community. We also implement strategy 2 (an improved store) in the San Juan and Utah
County communities for comparison.

### Erect a new store

This strategy assumes that the nutrition environment would benefit from a new
store located in a place that presently has low grocery store access. To examine
the potential for this strategy to improve access to nutrition in each
community, we calculate the change in destination choice log-sum when a new
store is added to the region in a location where access is currently poor. The
new store is a full-service grocery store with a number of registers equal to
the mean of other grocery stores in the community, and NEMS availability score,
NEMS cost score, and market basket cost equivalent to the 75$^{th}$ percentile
for the community. Thus the store is expected to be better-than-average quality
as perceived by the residents of the community. The location for this new store
is at 4100 S and 2700 W in West Valley City.

### Improve an existing store

This strategy assumes that existing stores are in locations that the community
values and can access, but that those stores may not have high availability of
quality goods. To examine the potential for this strategy to improve access to
nutrition, we improve the attributes of an existing dollar store in the
community so that it has the size, prices, and availability of goods as a
full-service grocery store. As above, we create a full-service grocery store
with a number of registers equal to the mean of other grocery stores in the
community, and NEMS availability score, NEMS cost score, and market basket cost
equivalent to the 75$^{th}$ percentile for the community. Thus the store is
expected to be better-than-average quality as perceived by the residents of the
community; the difference from the previous scenario is that the improved store
takes the place of an existing convenience store or dollar store.

The improved stores are at the following locations in each community:

-   An ethnic store near 2700 W 3500 S in West Valley City (Salt Lake)
-   A small grocery store in Santaquin (Utah County)
-   A dollar store in Blanding (San Juan County)

### Improve transit and non-motorized transport

This strategy assumes that people cannot easily travel to existing stores
because they cannot or do not drive for a variety of reasons, and that the
public and active transport networks provide an insufficient level of
service. To examine the potential for this strategy to improve access to
nutrition, we improve the travel time costs in the Salt Lake community for
non-motorized and public transportation in the region and calculate the change
in destination choice log-sum.

For active transportation, the lack of pedestrian facilities across and
alongside roads both in reality and in the OpenStreetMap dataset may
substantially increase measured walk distances and times. In this scenario, we
replace the times measured from OpenStreetMap using R5 with an idealized
Euclidean distance function, $$
t_{\mathrm{walk}} = \frac{\sqrt{2} * d'_{ij}}{v_{\mathrm{walk}}}
$$ {#eq-newwalk} where $d'_{ij}$ is the Euclidean (straight-line) distance
between $i$ and $j$ and $v_{ij}$ is an average walking speed equivalent to 3.5
feet per second [@fitzpatrick2006]. The distance is multiplied by the square
root of 2 to reflect the Manhattan distance (along a gridded street system). We
retain the cap on walking distance at 10 kilometers. Though this distance may
radically understate the real walking distance, we are trying to create an
idealized scenario of effectively frictionless active transport. For public
transit, we assume that the frequency of service is such that all transfer and
initial wait times are at most 5 minutes, and that no person must walk more than
10 minutes to access their first public transport service.

Travel times are improved in this way for all block group --- store pairs in the
west Salt Lake Valley community.

### Grocery delivery services

This strategy assumes that the most effective way to bring quality nutrition to an 
inaccessible is not to change the distribution of stores or transport, but to
support and enable delivery services that will bring the nutrition to individuals.
To model this scenario, we assume that 30% of the full-service grocery stores in 
the west Salt Lake Valley community offer a delivery service. This raises the
price of the market basket by $10, but reduces the travel time on all modes to 0
if the store is within 5 kilometers.
All of these assumptions are imperfect: a local grocery chain presently offers
services at a $7 delivery fee, which is not paid by all customers; additionally, 
delivery services require that customers expend time placing online orders; and 
service providers typically place a long-distance charge rather than a service area
limit. Regardless, these assumptions will work for the purpose of this exercise.

## Scenario Results

```{r loadbase}
tar_load(bg)
tar_load(bg_acs)
sl_base <- tar_read(sl_access)
tar_load(mkt_betas)
```

Using the methodology described above, we recalculated the destination choice
log-sum value for each block group under each scenario, and compared the change
in accessibility resulting from the improvement.

```{r s1results}
#| label: fig-s1results
#| fig-cap: Estimated per-household benefits of adding new full-service store in west Salt Lake Valley
#| out-width: "6in"
s1_altd <- tar_read(s1_access)

s1cost <- bg |>  select(id = GEOID) |> filter(substr(id, 3, 5) == "035") |> 
  left_join(bg_acs, by = c("id" = "geoid")) |> 
  sf::st_transform(4326) |> 
  left_join(sl_base |> rename(base = dclogsum)) |> 
  left_join(s1_altd |> rename(alt = dclogsum)) |> 
  mutate(diff = (base - alt) / mkt_betas["SaltLake.market"])
pal <- wesanderson::wes_palette("Zissou1", 100, type = "continuous")

ggplot(s1cost, aes(fill = diff)) + 
  ggspatial::annotation_map_tile(type = "cartolight", zoom = 12) +
  geom_sf(lwd = 0, alpha = 0.5) +
  scale_x_continuous(limits = c(-112.11, -111.7)) + 
  scale_y_continuous(limits = c(40.5, 40.8)) + 
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_gradientn("Benefit [$]", colours = pal)
```

@fig-s1results shows the geographic distribution of benefits associated with
locating a new store at a site in the Salt Lake community. The benefits are
largest immediately next to the new store, where they exceed
$`r floor(max(s1cost$diff, na.rm = TRUE))`$ for each household each time the
household makes a trip to a grocery store.

```{r s2results_sl}
#| label: fig-s2results
#| fig-cap: Estimated per-household benefits of improving an existing store in
#|   west Salt Lake Valley
#| out-width: "6in"
s2_altd <- tar_read(s2_access)

s2cost <- bg |>  select(id = GEOID) |> filter(substr(id, 3, 5) == "035") |> 
  left_join(bg_acs, by = c("id" = "geoid")) |> 
  sf::st_transform(4326) |> 
  left_join(sl_base |> rename(base = dclogsum)) |> 
  left_join(s2_altd |> rename(alt = dclogsum)) |> 
  mutate(diff = (base - alt) / mkt_betas["SaltLake.market"])
pal <- wesanderson::wes_palette("Zissou1", 100, type = "continuous")

ggplot(s2cost, aes(fill = diff)) + 
  ggspatial::annotation_map_tile(type = "cartolight", zoom = 12) +
  geom_sf(lwd = 0, alpha = 0.5) +
  scale_x_continuous(limits = c(-112.11, -111.7)) + 
  scale_y_continuous(limits = c(40.5, 40.8)) + 
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_gradientn("Benefit [$]", colours = pal)
```

@fig-s2results shows the results of the scenario improving an existing store in
the Salt Lake community. Compared to the results of the new store scenario, the
scale of the benefits are not as substantial (a maximum per-household-trip
benefit of less than \$1), and seem to not cover quite as large a geographic
region. @fig-s2sjut shows the results of improving a store in Utah and San Juan
Counties. As in Salt Lake, the benefits are most strongly concentrated in the
immediate vicinity of the improved store. One interesting observation ---
especially in Utah County --- is that the improvements are felt more strongly in
the block groups near the improved store that have lower availability of other
options. The block group in Utah County directly containing the improvement sees
a per-household-trip benefit well over \$2, considerably more than the maximum
benefit in Salt Lake. This is intuitive, as the improvement of a store matters
less if the stores close to you are already sufficient.

```{r s2results-utsj}
#| label: fig-s2sjut
#| fig-cap: Estimated per-household benefits of improving an existing store in Utah and San Juan Counties.
#| layout-ncol: 2
#| fig-subcap: 
#|   - Utah County (in Santaquin)
#|   - San Juan County (in Blanding)
#| out-width: "4in"
tar_load(wf_access)
tar_load(ru_access)
tar_load(s2_access_ut)
tar_load(s2_access_sj)

pal <- wesanderson::wes_palette("Zissou1", 100, type = "continuous")

s2cost_ut <- bg |>  select(id = GEOID) |> filter(substr(id, 3, 5) == "049") |> 
  left_join(bg_acs, by = c("id" = "geoid")) |> 
  sf::st_transform(4326) |> 
  left_join(wf_access |> rename(base = dclogsum)) |> 
  left_join(s2_access_ut |> rename(alt = dclogsum)) |> 
  mutate(diff = (base - alt) / mkt_betas["Utah.market"])

ggplot(s2cost_ut, aes(fill = diff)) + 
  ggspatial::annotation_map_tile(type = "cartolight", zoom = 12) +
  scale_x_continuous(limits = c(-112.1, -111.5)) + 
  scale_y_continuous(limits = c(39.9, 40.48)) + 
  geom_sf(lwd = 0, alpha = 0.5) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_gradientn("Benefit [$]", colours = pal)


s2cost_sj <- bg |>  select(id = GEOID) |> filter(substr(id, 3, 5) == "037") |> 
  left_join(bg_acs, by = c("id" = "geoid")) |> 
  sf::st_transform(4326) |> 
  left_join(ru_access |> rename(base = dclogsum)) |> 
  left_join(s2_access_sj |> rename(alt = dclogsum)) |> 
  mutate(diff = (base - alt) / mkt_betas["SanJuan.market"])


ggplot(s2cost_sj, aes(fill = diff)) + 
  ggspatial::annotation_map_tile(type = "cartolight", zoom = 10) +
  geom_sf(lwd = 0, alpha = 0.5) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_gradientn("Benefit [$]", colours = pal)

```

The results of the third scenario, improving the access of non-motorized and
public transit access to stores, are shown in @fig-s3results. This benefit is
spread over a larger area, and is concentrated on the 35 MAX bus rapid transit
corridor where the improvement in walk access time to transit couples with high
frequency transit service to large grocery stores on the corridor. The
per-household-trip benefit is very small however, with a maximum benefit on the
order of \$0.25.

```{r s3results}
#| label: fig-s3results
#| fig-cap: Estimated per-household benefits of improving non-motorized and public transport access to groceries in Salt Lake Valley.
#| out-width: "6in"
s3_altd <- tar_read(s3_access)

s3cost <- bg |>  select(id = GEOID) |> filter(substr(id, 3, 5) == "035") |> 
  left_join(bg_acs, by = c("id" = "geoid")) |> 
  sf::st_transform(4326) |> 
  left_join(sl_base |> rename(base = dclogsum)) |> 
  left_join(s3_altd |> rename(alt = dclogsum)) |> 
  mutate(diff = (base - alt) / mkt_betas["SaltLake.market"])

pal <- wesanderson::wes_palette("Zissou1", 100, type = "continuous")
tar_load(gtfs_shape)
gtfs_shape <- sf::st_simplify(gtfs_shape, dTolerance = 10)

ggplot(s3cost, aes(fill = diff)) + 
  ggspatial::annotation_map_tile(type = "cartolight", zoom = 12) +
  geom_sf(lwd = 0, alpha = 0.5) +
  geom_sf(data = gtfs_shape |> filter(mode != "BRT"), aes(color = mode)) +
  scale_x_continuous(limits = c(-112.11, -111.7)) + 
  scale_y_continuous(limits = c(40.5, 40.8)) + 
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_gradientn("Benefit [$]", colours = pal)
  scale_color_manual("Transit", values = c("gray80", "black", "gray50"))
```

```{r s4results}
#| label: fig-s4results
#| fig-cap: Estimated per-household benefits of grocery deliveries in Salt Lake Valley.
#| out-width: "6in"
s4_altd <- tar_read(s4_access)

s4cost <- bg |>  select(id = GEOID) |> filter(substr(id, 3, 5) == "035") |> 
  left_join(bg_acs, by = c("id" = "geoid")) |> 
  sf::st_transform(4326) |> 
  left_join(sl_base |> rename(base = dclogsum)) |> 
  left_join(s4_altd |> rename(alt = dclogsum)) |> 
  mutate(diff = (base - alt) / mkt_betas["SaltLake.market"])

pal <- wesanderson::wes_palette("Zissou1", 100, type = "continuous")

ggplot(s4cost, aes(fill = diff)) + 
  ggspatial::annotation_map_tile(type = "cartolight", zoom = 12) +
  geom_sf(lwd = 0, alpha = 0.5) +
  scale_x_continuous(limits = c(-112.11, -111.7)) + 
  scale_y_continuous(limits = c(40.5, 40.8)) + 
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_gradientn("Benefit [$]", colours = pal)
```


```{r scenarios}
#| label: tbl-scenarios
#| tbl-cap: Scenario Benefits
df <- list(
  "New Store" = s1cost,
  "Salt Lake Valley" = s2cost,
  "Utah County" = s2cost_ut,
  "San Juan County" = s2cost_sj,
  "Improved Transport" = s3cost,
  "Delivery Services" = s4cost
) |> 
  bind_rows(.id = "Scenario") |> 
  sf::st_set_geometry(NULL) |> 
  mutate(
    hh = households * diff,
    nw = (100 - white)/100 * households * diff,
    li = households * lowincome/100 * diff
  ) |> 
  summarise(
    across(hh:li, ~scales::dollar(sum(.x, na.rm = TRUE) / mkt_betas["SaltLake.market"] * -1)),
    .by = Scenario
  ) 

if(!knitr::pandoc_to("docx")){
  kbl(df, booktabs = TRUE, col.names = c("Scenario", "Households", "Non-white", "Low-Income"),
      digits = 0) |> 
  add_header_above(c(" " = 1, "Weighted by" = 3)) |> 
  group_rows(group_label = "Improved Store", start_row = 2, end_row = 4) |> 
  kable_styling()
}
```

Although comparing the geographic distribution of benefits is helpful, the
aggregate benefit is more likely to guide policy. Additionally, the aggregate
benefits can be weighted in different ways to understand the effects of the
various policies on different populations. @tbl-scenarios presents the aggregate
benefit from each of the three scenarios (and the result of the second scenario
in all three communities). The Households column multiplies the difference in
destination choice log-sum at each block group by the number of households in
that block group, while the non-white and low-income columns weight the
difference by the share of non-white individuals and low-income households
respectively. All demographic data comes from the American Community Survey
5-year aggregations.

These alternate weighting schemes help to illustrate the potential equity of the
benefit distribution should each scenario be pursued. The new store scenario in
the Salt Lake community, for example, has a total benefit of approximately \$5.5
million. Of this amount, somewhat less than half the benefits go to non-white
individuals and less than one-fifth to low-income households. These ratios are
more or less the same for the store improvement scenario and the improved
transport scenario in the Salt Lake Valley. The Utah County store improvement,
on the other hand, has a somewhat higher proportion of benefits going to
low-income households relative to non-white households. This is a reflection of
the lower minority population in southern Utah County vis a vis west Salt Lake
valley, but is nonetheless a metric that program evaluators might pay attention
to.

Overall, the improved store brings more than twice the benefits of improving
non-automobile transportation, and the new store more than five times the
benefits. Understanding the costs of these various alternatives is outside the
scope of this research, but the level of infrastructure investment required to
increase transportation level of service to that constructed in the simulation
is likely an order of magnitude higher than the cost of a single new grocery
store. It should also be acknowledged that improving non-automobile
infrastructure and services would have benefits beyond just grocery store trips
that we do not attempt to enumerate here. It is also not clear whether the level
of improvement simulated in the second scenario could be accomplished within the
envelope of the existing store; it may be that such improvements would meet or
exceed the cost of building a new store on a new site, along with stocking,
staffing, and operating the store.
